Identifying Inter-Genomic Repeats Using Fast, Approximate Methods for Betweenness Centrality

### 2015-05-18 manuscript/manuscript.tex
Small edits.

\subsection*{Evaluation}
We evaluate the effect of using approximation algorithms for betweenness centrality on finding inter-genome repeats using the simulated datasets described above.
%We use the notion of Sensitivity, specificity, and precision to define the quality of repeat detection.
A good repeat detection strategy should have high sensitivity, specificity, and precision. Sensitivity is the measure of proportion of actual repeats which are correctly identified by the algorithm. Specificity is the proportion of non-repeats correctly identified by the algorithm. Precision is the proportion of repeats identified that are correct. High sensitivity is desired because we want to remove as many repeats from the graph. High specificity is desired because we do not want to remove non-repeats which are incorrectly identified as repeats by algorithm. High precision helps us determine the usefulness of our methods. To calculate sensitivity, specificity, and precision, we define following terms:
\begin{itemize}
\item \textbf{True positives(TP):} Number of nodes marked as a repeat that is found within at least 2 genomes
\item \textbf{False positives(FP):} Number of nodes marked as a repeat that is only found within a single genome
\item \textbf{True negatives(TN):} Number of nodes marked as a non-repeat that is only found within a single genome
\item \textbf{False negatives(FN):} Number of nodes marked as a non-repeat that is found within at least 2 genomes
\end{itemize}

@@

For evaluating the effect of node sampling on inter-genomic repeat detection, we sample between 10 to 10,000 nodes per trial.
For evaluating path sampling, we select epsilon parameters between 0.01 and 0.20.
The result of each parameter is averaged over 10 trials.

@@

\subsection*{Implementation}
We investigate the performance of approximate betweenness centrality algorithms on the metagenomic graphs to find inter-genome repeats. We carried out experiments on a 64-bit Linux machine with 12 CPUs and 50 GB memory. We used the implementations provided by \textsc{NetworKit}\cite{networkit} library. It implements effecient graph algorithms and parallelize most of them. It provides implementation for variety of centrality measures. We used the implementation of approximate betweenness centrality. For our experiments, we use the parallelized implementation for the algorithm by Matteo et al.\cite{matteo} and serial implementation for the algorithm by Geisberger et al.\cite{sanders}.

@@

The advantage of using the node sampling algorithm is that it appears to scale linearly in practice.
Thus, if runtime is a concern, small increasing subsets of nodes can be sampled and the runtime of a larger sample can be extrapolated.
We can pair this information with how the subset of nodes marked as inter-genomic repeats changes as we increase the sample size.
If the same set of nodes are being marked as inter-genomic repeats, then the algorithm can be terminated before sampling a larger set of nodes.
Conversely, if there is high variability between nodes marked as repeats in subsequent samplings, then a larger set of nodes should be sampled until the results stabilize.
This is the main advantage of the shortest path sampling approach.
While the shortest path sampling approach takes a fair bit longer in our evaluation then sampling a small number of nodes, it has a probabilistic guarantee associated with the betweenness centrality computation.

>>>>

\subsection*{Evaluation}
%We evaluate the effect of using approximation algorithms for betweenness centrality on finding inter-genome repeats using the simulated datasets described above.
%We use the notion of Sensitivity, specificity, and precision to define the quality of repeat detection.
A good repeat detection strategy should have high sensitivity, specificity, and precision. Sensitivity is the measure of proportion of actual repeats which are correctly identified by the algorithm. Specificity is the proportion of non-repeats correctly identified by the algorithm. Precision is the proportion of repeats identified that are correct. High sensitivity is desired because we want to remove as many repeats from the graph. High specificity is desired because we do not want to remove non-repeats which are incorrectly identified as repeats by algorithm. High precision helps us determine the usefulness of our methods. To calculate sensitivity, specificity, and precision, we define following terms:
\begin{itemize}
\item \textbf{True positives(TP):} Number of nodes marked as a repeat that is found within at least 2 genomes
\item \textbf{False positives(FP):} Number of nodes marked as a repeat that is only found within a single genome
\item \textbf{True negatives(TN):} Number of nodes marked as a non-repeat that is only found within a single genome
\item \textbf{False negatives(FN):} Number of nodes marked as a non-repeat that is found within at least 2 genomes
\end{itemize}

@@

We evaluate the two approaches for calculating betweenness centrality based on sampling nodes and sampling shortest paths using the above measures.
For evaluating the effect of sampling nodes on inter-genomic repeat detection, we sample between 10 to 10,000 nodes per trial.
For evaluating shortest paths sampling, we select epsilon parameters between 0.01 and 0.20.
The result of each parameter is averaged over 10 trials.

@@

\subsection*{Implementation}
We investigate the performance of approximate betweenness centrality algorithms on the metagenomic assembly graphs using the implementations provided by \textsc{NetworKit}\cite{networkit} library. \textsc{NetworKit} contains efficient and parallel implementations of a collection of graph algorithms.
It provides implementations for various centrality measures, but here we only use the implementations related to betweenness centrality. For our experiments, we use the parallelized implementation for the algorithm by Matteo et al.\cite{matteo} and serial implementation for the algorithm by Geisberger et al.\cite{sanders}.
We carried out experiments on a 64-bit Linux machine with 12 CPUs and 50 GB memory.

@@

The advantage of calculating betweenness centrality using node sampling is that it appears to scale linearly in practice.
Thus, if runtime is a concern, small increasing subsets of nodes can be sampled and the runtime of a larger sample can be extrapolated.
We can pair this information with how the subset of nodes marked as inter-genomic repeats changes as we increase the sample size.
If the same set of nodes are being marked as inter-genomic repeats, then the algorithm can be terminated before sampling a larger set of nodes.
Conversely, if there is high variability between nodes marked as repeats in subsequent samplings, then a larger set of nodes should be sampled until the results stabilize.
This is the main advantage of the shortest path sampling approach.
While the shortest path sampling approach takes a fair bit longer in our evaluation then sampling a small number of nodes, it has a probabilistic guarantee associated with the betweenness centrality computation.
